---
title: "Introduction and System Setup"
---

## Overview

This two-day workshop will introduce you to the basics of working with telemetry data in R.

We will providing a data set to use throughout the workshop, but we encourage you to bring along your own data for the hackathon session (Friday AM).

### The workshop will include:

-   The basics of spatial data in R
-   How to read in, clean, and QA your telemetry data
-   How to create useful summaries of your data
-   Data visualization techniques for spatial and telemetry data (graphs and maps)
-   How to use the `bcmaps` and `bcdata` packages to get vector and raster data from official BC government sources.
-   How to perform spatial operations to compile and generate landcape covariates
-   How to extract spatial information for telemetry and random location points

### The workshop will not include:

-   Methodological details on technical telemetry parameters i.e. DOP estimates

-   Advanced modeling techniques

### Agenda

#### Day 1

- Setup and troubleshooting (8:30 - 9:00)
- Working with spatial data in R (9:00 - 10:30)
- **BREAK** (10:30 - 11:00)
- Telemetry data in R; clean, QA, and prepare data (Part I; 11:00 - 12:30)
- **LUNCH** (12:30 - 1:30)
- Telemetry data in R; clean, QA, and prepare data (Part II; 1:30 - 3:00)
- **BREAK** (3:00 - 3:30)
- Visualizing spatial data in R (3:30 - 4:30)

#### Day 2

- Setup and troubleshooting (8:30 - 9:00)
- Retrieving spatial data from the B.C. Data Catalogue (9:00 - 10:30)
- **BREAK** (10:30 - 11:00)
- Getting and working with raster data (Part I; 11:00 - 12:30)
- **LUNCH** (12:30 - 1:30)
- Preparing  base data for telemetry analysis (Part II; 1:30 - 3:00)
- **BREAK** (3:00 - 3:30)
- Preparing data for Resource Selection Function (RSF) analysis (3:30 - 4:30)

## Computing requirements

You will require the following software installed and configured for the workshop. Please have this set up and ready to go **before** we start.

You will need:

-   A laptop computer, preferably with administrative privileges
-   [R and RStudio](#install-r-and-rstudio)
-   [Several R packages](#packages)
-   [QGIS](qgis) (optional)

### Install R and RStudio {#install-r-and-rstudio}

You will need:

-   R version \>= 4.2.0
-   RStudio \>= 2023.03.1

#### Install R

Download and install R for your operating system from <https://cloud.r-project.org/>.

##### Windows with no admin rights:

If you do not have administrator rights, the installer will default to installing somewhere in your user folder (e.g., `C:/Users/username/AppData/Local/Programs/`). If you prefer, you can change the location to another folder that you have write access to, just make sure it is on your `C:/` drive.

#### Install R Studio

Download and install RStudio Desktop from <https://posit.co/download/rstudio-desktop/>. This page should automatically offer you the version suitable for your operating system, but you can scroll down to find versions for all operating systems.

##### Windows with admin rights:

Download the `.zip` archive for Windows under "Zip/Tarballs". Create a folder called `RStudio` in a location on your `C:/` drive, where you have write access (e.g. `C:/Users/username/AppData/Local/Programs/RStudio`), and extract the zip file into this folder. Find the RStudio program in this folder: it is named `rstudio.exe`, but the file extension will typically be hidden, so look for `rstudio`. Right-click this file to create a shortcut and drag it to your desktop/task bar. Use the shortcut to open RStudio.

### Install packages {#packages}

In R, install the necessary packages by running:

```{r}
#| eval: false
install.packages(
  c("tidyverse", "sf", "terra", "mapview", "bcdata", "bcmaps", "readxl", "ggplot2", "usethis")
)
```

### QGIS (optional) {#qgis}

QGIS is a free and open-source geographic information system (GIS) that allows you to create, edit, visualize, analyze and publish geospatial information.

Download and install from the [QGIS website](https://qgis.org/en/site/forusers/download.html).

## Workshop materials and data

We have created an empty RStudio project containing the workshop data.

The easiest way to get started to run:

```{r}
#| eval: false
usethis::use_course("https://tinyurl.com/3dmxtxaz")
```

This will download the workshop materials, save them in a logical spot on your computer, 
and open the project in a new RStudio session.

Alternatively, you can download the zip file from [here](https://github.com/ninoxconsulting/r-telemetry-workshop-nov-2023/raw/main/r-telemetry-workshop.zip), decompress the zip file somewhere convenient on your C:/ drive. Then, navigate to the newly created folder (called `r-telemetry-workshop`), 
and double click on the `r-telemetry-workshop.Rproj` file to open the project in RStudio.
