---
title: "Interactive and animated maps"
---

## Animated maps

We can visualize movement by animating their tracks, using the `{gganimate}` 
package.

```{r}
#| message: false
library(dplyr)
library(ggplot2)
library(sf)
library(gganimate)
library(ggspatial)
```

Let's start with the map we made earlier of the paths of two animals, 
faceted by year:

```{r}
scott <- read_sf("clean_data/scott_herd_subset.gpkg")

scott <- cbind(st_coordinates(scott), scott)

top_animals <- c("SC_car171", "SC_car168")

mvmt <- scott |> filter(animal.id %in% top_animals) |> 
  mutate(doy = lubridate::yday(date2))


p <- ggplot(mvmt) + 
  geom_path(aes(x = X, y = Y, colour = doy)) + 
  facet_grid(animal.id ~ year) +
  scale_color_viridis_c() + 
  coord_sf() + 
  labs(colour = "Day of Year") + 
  theme_bw() + 
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    legend.position = "bottom",
    legend.key.width = unit(3, "lines")
  )
```

We then add a `transition` to the plot that reveals each step along the path, 
and create an animated gif:

```{r}
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 6

p_animate <- p + 
  transition_reveal(along = doy) + 
  labs(title = "Day of Year: {round(frame_along)}")

animate(
  p_animate, 
  width = 10, 
  height = 6, 
  units = "in", 
  res = 72, 
  fps = 10, 
  nframes = 300
)
```

## Interactive maps with `{Leaflet}`

Let's use the file that has the caribou locations with the attributes from 
the base layers

```{r}
library(dplyr)
library(leaflet)

all_pts <- read_sf("clean_data/allpts_att.gpkg") |> 
  st_transform(4326)

caribou <- filter(all_pts, !is.na(animal.id))
background <- filter(all_pts, is.na(animal.id))

animal_pal <- colorFactor(
  palette = "viridis",
  domain = unique(caribou$animal.id)
)

leaflet(caribou) |> 
  addProviderTiles(providers$Esri.NatGeoWorldMap) |> 
  addCircleMarkers(
    data = background, 
    color = "grey", 
    radius = 1
    ) |> 
  addCircleMarkers(
    color = ~ animal_pal(caribou$animal.id),
    weight = 1,
    radius = ~ elevation / 300,
    popup = ~ paste0(
      "Animal: ", animal.id, "<br>",
      "Elevation: ", round(elevation)
    )
  ) |> 
  addLegend(
    "bottomleft", 
    pal = animal_pal,
    values = ~ caribou$animal.id,
    title = "Animal ID"
  )
```

